---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# astgrepr

<!-- badges: start -->
[![R-CMD-check](https://github.com/etiennebacher/astgrepr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/etiennebacher/astgrepr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`astgrepr` provides R bindings to the [ast-grep](https://ast-grep.github.io/) 
Rust crate. `ast-grep` is a tool to parse the abstract syntax tree (AST) of some
code and to perform search and rewrite of code. This is extremely useful to build
linters, stylers, and perform a lot of code analysis.

Since `astgrepr` can be used as a low-level foundation for other tools (such as
linters), it doesn't rely on any R dependencies (but has Rust dependencies 
brought in by `ast-grep`).

```{r}

library(astgrepr)

src <- "library(tidyverse)
    x <- rnorm(100, mean = 2)
    any(duplicated(y))
    plot(x)
    any(duplicated(x))"

node <- src |> 
  tree_new() |> 
  tree_root()

# get everything inside rnorm()
node |> 
  node_find(pattern = "rnorm($$$A)") |> 
  node_get_multiple_matches("A") |> 
  node_text_all()

# find occurrences of any(duplicated())
node |> 
  node_find_all(pattern = "any(duplicated($A))") |> 
  node_text_all()

```

