---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(astgrepr)
```

The original [`ast-grep`](https://ast-grep.github.io/guide/pattern-syntax.html) library has a great documentation where you can learn more about pattern syntax and writing advanced rules. This vignette will focus mostly on the design implementation in R so that you understand how the package can be used.

Functions in `astgrepr` can be classed in two categories: those that should be applied on a single node, and those that should be applied on multiple nodes. But before jumping to this distinction, let's create a simple code example and generate a tree, which is essential in order to use `astgrepr` functionalities:

```{r}
src <- "x <- rnorm(100, mean = 2)
any(duplicated(y))
plot(x)
any(duplicated(x))"

root <- src |>
  tree_new() |>
  tree_root()
```

## Single-node functions

## Multi-node functions